---
title: "lab3"
format: pdf
editor: visual
---

```{r, warning=FALSE}
library(opendatatoronto)
library(tidyverse)
library(stringr)
library(janitor)
library(lubridate)
library(ggrepel)
library(dplyr)
library(bayestestR)
```

# Question 1

$Y|\theta$ \~ $Bin(n,\theta)$\
likelihood function: $L(y,\theta,n)=\binom{n}{y}\theta^y(1-\theta)^{n-y}$\
log-likelihood function: $l(y,\theta,n)=log(L)=log(\binom{n}{y}\theta^y(1-\theta)^{n-y})$ $=log\binom{n}{y}+log(\theta^y)+log((1-\theta)^{n-y})$ $=log\binom{n}{y}+ylog(\theta)+(n-y)log(1-\theta)$\
take derivative respect to $\theta$ and set to 0:\
$\frac{dl}{d\theta}=\frac{y}{\theta}-\frac{n-y}{1-\theta}=0$\
$\hat{\theta}=\frac{y}{n}=\frac{118}{129}=0.91$\
Fisher Information:\
$I(\hat{\theta})=-E[l''(\hat{\theta})]=\frac{118}{0.91^2}+\frac{11}{(1-0.91)^2}=1500.52$\
95% confidence interval:\
$(\hat{\theta}-1.96\frac{1}{\sqrt{I(\hat{\theta})}},\hat{\theta}+1.96\frac{1}{\sqrt{I(\hat{\theta})}})$\
$(0.91-0.05,0.91+0.05)$\
$(0.86, 0.96)$

# Question 2

$\theta$ \~$beta(1,1)$\
$p(\theta)=\frac{\gamma(1+1)}{\gamma(1)\gamma(1)}\theta^{1-1}(1-\theta)^{1-1}=1$\
$p(\theta|y)=\frac{p(y|\theta)p(\theta)}{p(y)}=\frac{p(y|\theta)p(\theta)}{\int p(y|\theta')d\theta'}$\
$=\frac{\binom{n}{y}\theta^y(1-\theta)^{n-y}}{\int_{0}{1}\binom{n}{y}\theta'^y(1-\theta')^{n-y}d\theta'}$\
$=\frac{1}{Z}\theta^y(1-\theta)^{n-y}$, where $Z = \frac{\gamma(y+1)\gamma(n-y+1)}{\gamma(n+2)}$\
the posterior distribution is :\
$\theta|y$\~$Beta(y+1,n-y+1)$\
The posterior mean:\
$E(θ|y)=\frac{y+1}{y+1+n-y+1}=\frac{y+1}{n+2}=\frac{118+1}{129+2}=0.91$\
95% Credible interval:

```{r}
posterior <- distribution_beta(1000, 119, 12)
ci_eti <- ci(posterior, method = "ETI")
ci_eti
```

# Question 3

$\theta$ \~$beta(10,10)$\
interpretation: my subjective beliefs about the parameter $\theta$ is distributed $beta(10,10)$.\
we assuming we know more amount of information as the prior used in Question 2, since the the prior in Question 2 is uniform, and contain no information.

# Question 4

black: The likelihood\
red,dashed: The prior distribution of prior beta(1,1)\
red,solid: The posteriors distribution of prior beta(1,1)\
blue, dashed: The prior distribution of prior beta(10,10)\
blue, dahsed: The posteriors distribution of prior beta(10,10)

```{r}
p = seq(from = 0, to =1, by = 0.001)
data = data.frame(theta=p,ptheta=dbinom(118, size = 129, p))
ggplot(data, aes(x=theta,y=ptheta))+geom_line()+stat_function(fun=dbeta, args=list(shape1=1, shape2=1),color="red",linetype = "dashed")+stat_function(fun=dbeta, args=list(shape1=119, shape2=12),color="red")+stat_function(fun=dbeta, args=list(shape1=10, shape2=10),color="blue",linetype = "dashed")+stat_function(fun=dbeta, args=list(shape1=128, shape2=21),color="blue")
```

# Question 5

A noninformative prior: has distribution which is flat over the entire real number line and contain no information, for example, uniform distribution.

A subjective/informative prior based on your best knowledge: has a distribution which containing specific, unambiguous information about the variable, for example, based on the distribution of the average improvement in success probability.
