---
title: "lab5"
date: today
date-format: "DD/MM/YY"
format: pdf
execute: 
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(rstan)
library(tidybayes)
library(here)
```

```{r}
kidiq <- read_rds(here("data","kidiq.RDS"))
kidiq
```

## Question 1

```{r}
boxplot(kid_score ~mom_hs , data = kidiq)
```

The first interest point is that the status of mother's completion of high school is a binary data 0 and 1. Based on the box plot, the children of mothers who completed high school scored higher than children of mothers who did not.

```{r}
ggplot(kidiq,aes(x=mom_iq,y=kid_score))+geom_point()+geom_smooth(method='lm')
```

The second interest point is that the kid score increases as the their mom's iq increases.

```{r}
ggplot(kidiq,aes(x=mom_age,y=kid_score))+geom_point()+geom_smooth(method='lm')
```

The third interest point is that the kid score do not have strong relationship with the their mom's age.

## Question 2

```{r}
y <- kidiq$kid_score
mu0 <- 80
sigma1 <- 0.1
```

```{r}
data1 <- list(y = y, 
             N = length(y), 
             mu0 = mu0,
             sigma0 = sigma1)
```

```{r}
fit1 <- stan(file = here("code/models/kids2.stan"),
            data = data1,
            chains = 3,
            iter = 500)
```

```{r}
fit1
```
The estimated coefficients of fit1 model is mu 80.06 and sigma 21.44. Comparing to fit model(mu 86.74 sigma 20.40), the mu value decreased and the sigma value slightly increased.

```{r}
dsamples <- fit1  |> 
  gather_draws(mu, sigma) # gather = long format
```

```{r}
dsamples |> 
  filter(.variable == "mu") |> 
  ggplot(aes(.value, color = "posterior")) + geom_density(size = 1) + 
  xlim(c(79, 81)) + 
  stat_function(fun = dnorm, 
        args = list(mean = mu0, 
                    sd = sigma1), 
        aes(colour = 'prior'), size = 1) +
  scale_color_manual(name = "", values = c("prior" = "red", "posterior" = "black")) + 
  ggtitle("Prior and posterior for mean test scores") + 
  xlab("score")
```

## Question 3

```{r}
X <- as.matrix(kidiq$mom_hs, ncol = 1) # force this to be a matrix
K <- 1

data <- list(y = y, N = length(y), 
             X =X, K = K)
fit2 <- stan(file = here("code/models/kids3.stan"),
            data = data, 
            iter = 1000)
```

```{r}
summary(lm(kid_score~mom_hs,data = kidiq))
```

(a) The coefficients of fit2 is alpha 77.773 beta\[1\] 11.47.\
    The coefficients of lm() model is (Intercept) 77.548 and mom_hs 11.771. Their value are very close to each other.

```{r}
pairs(fit2, pars = c("alpha", "beta[1]"))
```

(b) There is a large variation in alpha and beta\[1\], it would make the sampling harder.

## Question 4

```{r}
center_scale <- function(x) {
    scale(x, scale = FALSE)
}
```

```{r}
X4 <- cbind(X,center_scale(kidiq$mom_iq)) # force this to be a matrix
K4 <- 2
data4 <- list(y = y, N = length(y), 
             X =X4, K = K4)
fit4 <- stan(file = here("code/models/kids3.stan"),
            data = data4, 
            iter = 1000)
```

```{r}
fit4
```

Fixed the mother completing high school status, For every 1 increase in IQ, the child's score increases by 0.57; Fixed the mother's IQ, Children of mothers who completed high school scored 5.63 higher than children of mothers who did not.

## Question 5

```{r}
summary(lm(kid_score~mom_hs+center_scale(mom_iq),data=kidiq))
```

The coefficients of fit4 is alpha 82.36 beta\[1\] 5.63 beta\[2\] 0.57.\
The coefficients of lm() model is (Intercept) 82.122 mom_hs 5.95 and center_scale(mom_iq) 0.564. Their value are very close to each other.

## Question 6

```{r}
IQ6 = 110- mean(kidiq$mom_iq)
```

```{r}
pivot_wider(spread_draws(fit4,alpha, beta[k], sigma),names_from = k,names_prefix = "beta",values_from = beta)
```

```{r}
fit4 |>
  spread_draws(alpha, beta[k], sigma) |> 
  pivot_wider(names_from = k,names_prefix = "beta",values_from = beta)|>
  mutate(nhs = alpha+0*beta1+IQ6*beta2, # no high school is just the intercept
          hs = alpha + 1*beta1 + IQ6*beta2) |> 
  select(nhs, hs) |> 
  pivot_longer(nhs:hs, names_to = "education", values_to = "estimated_score") |> 
  ggplot(aes(y = education, x = estimated_score)) +
  stat_halfeye() + 
  theme_bw() + 
  ggtitle("Posterior estimates of scores by education level of mother")
  
```

## Question 7

```{r}
IQ7 = 95-mean(kidiq$mom_iq)
post_samples4 <- extract(fit4)
prediction <- post_samples4[["alpha"]]+post_samples4[["beta"]][,1]*1+post_samples4[["beta"]][,2]*IQ7
y4 <- rnorm(mean= prediction, sd = post_samples4[['sigma']],n=2000)
hist(y4)
```
