---
title: "Lab 6"
format: pdf
editor: visual
---

```{r message=FALSE}
library(tidyverse)
library(here)
# for bayes stuff
library(rstan)
library(bayesplot) 
library(loo) 
library(tidybayes) 
```

```{r}
ds <- read_rds(here("data","births_2017_sample.RDS"))
head(ds)
```

```{r}
ds <- ds %>% 
  rename(birthweight = dbwt, gest = combgest) %>% 
  mutate(preterm = ifelse(gest<32, "Y", "N")) %>% 
  filter(ilive=="Y",gest< 99, birthweight<9.999)
```

## Question 1

```{r}
ggplot(ds,aes(x=log(gest),y=log(birthweight),color = preterm))+geom_point()+geom_smooth(method = lm)
```

The first interest point is that as gestational age increases, the infant's weight increases, and in the preterm period, the baby's body weight increases more rapidly.

```{r}
ggplot(ds,aes(x=birthweight,y=mager))+geom_point()
```

The second interest point is that there does not seem to be a strong relationship between the baby's weight and the mother's age.

```{r}
ggplot(ds,aes(x=log(gest),y=log(birthweight)))+geom_point()+geom_smooth(method=lm,aes(color=sex))
```

The third interest point is that as the gestational age increases, the birth weight increase for both baby boys and baby girls, and increase rate is slightly higher in baby girls than baby boys.

## Question 2

```{r warning=FALSE}
set.seed(100)
nsims <- 1000
sigma <- rnorm(nsims, 0, 1)
beta1 <- rnorm(nsims, 0, 1)
beta2 <- rnorm(nsims, 0, 1)

dsims <- tibble(log_gest_c = (log(ds$gest)-mean(log(ds$gest)))/sd(log(ds$gest)))

for(i in 1:nsims){
  this_mu <- beta1[i] + beta2[i]*dsims$log_gest_c 
  dsims[paste0(i)] <- this_mu + rnorm(nrow(dsims), 0, sigma[i])
}

dsl <- dsims %>% 
  pivot_longer(`1`:`1000`, names_to = "sim", values_to = "sim_weight")

dsl %>% 
  ggplot(aes(sim_weight)) + geom_histogram(aes(y = ..density..), bins = 20, fill = "turquoise", color = "black")
```

```{r warning=FALSE}
dsl10 <- dsims %>% 
  pivot_longer(`1`:`10`, names_to = "sim", values_to = "sim_weight")
ggplot(dsl10,aes(x=log_gest_c,y=sim_weight,color=sim))+geom_point()+geom_smooth()+theme_bw()
```

## Run the model

Now we're going to run Model 1 in Stan. The stan code is in the `code/models` folder.

First, get our data into right form for input into stan.

```{r}
ds$log_weight <- log(ds$birthweight)
ds$log_gest_c <- (log(ds$gest) - mean(log(ds$gest)))/sd(log(ds$gest))

# put into a list
stan_data <- list(N = nrow(ds),
                  log_weight = ds$log_weight,
                  log_gest = ds$log_gest_c)
```

Now fit the model

```{r}
mod1 <- stan(data = stan_data, 
             file = here("code/models/simple_weight.stan"),
             iter = 500,
             seed = 243)
```

```{r}
summary(mod1)$summary[c("beta[1]", "beta[2]", "sigma"),]
```

## Question 3

```{r}
est_37 <- (log(37)-mean(log(ds$gest)))/sd(log(ds$gest))
est_beta1 <- 1.1624783
est_beta2 <- 0.1437529
exp(est_beta1+est_beta2*est_37)
```

## Question 4

```{r warning = FALSE}
stan_data_mod2 <- list(N = nrow(ds),
                  log_weight = ds$log_weight,
                  log_gest = ds$log_gest_c,
                  preterm = ifelse(ds$preterm=="Y",1,0))

mod2 <- stan(data = stan_data_mod2, 
             file ="Lab6mod2.stan",
             iter = 500,
             seed = 243)
```

```{r}
summary(mod2)$summary[c("beta[1]", "beta[2]","beta[3]", "beta[4]", "sigma"),]
```

## Question 5

```{r}
load(here("output", "mod2.Rda"))
summary(mod2)$summary[c(paste0("beta[", 1:4, "]"), "sigma"),]
```

Based on the summary, the coefficients are similar after flipping beta\[2\] and beta\[3\].

## Question 6

```{r}
N = nrow(ds)
yrep2 <- extract(mod2)[['log_weight_rep']]
samp2 <- sample(nrow(yrep2),100)
# first, get into a tibble
rownames(yrep2) <- 1:nrow(yrep2)
dr <- as_tibble(t(yrep2))
dr <- dr %>% bind_cols(i = 1:N, log_weight_obs = log(ds$birthweight))

# turn into long format; easier to plot
dr <- dr %>% 
  pivot_longer(-(i:log_weight_obs), names_to = "sim", values_to ="y_rep")

# filter to just include 100 draws and plot!
dr %>% 
  filter(sim %in% samp2) %>% 
  ggplot(aes(y_rep, group = sim)) + 
  geom_density(alpha = 0.2, aes(color = "y_rep")) + 
  geom_density(data = ds %>% mutate(sim = 1), 
               aes(x = log(birthweight), col = "y")) + 
  scale_color_manual(name = "", 
                     values = c("y" = "darkblue", 
                                "y_rep" = "lightblue")) + 
  ggtitle("Distribution of observed and replicated birthweights") + 
  theme_bw(base_size = 16)

```

## Question 7

```{r}
yrep1 <- extract(mod1)[["log_weight_rep"]]
```

```{r}
y <- ds$log_weight
t_y <- mean(y<=log(2.5))
t_y_rep <- sapply(1:nrow(yrep1), function(i) mean(yrep1[i,]<=log(2.5)))
t_y_rep_2 <- sapply(1:nrow(yrep2), function(i) mean(yrep2[i,]<=log(2.5)))
```

```{r}
ggplot(data = as_tibble(t_y_rep), aes(value)) + 
    geom_histogram(aes(fill = "replicated")) + 
    geom_vline(aes(xintercept = t_y, color = "observed"), lwd = 1.5) + 
  ggtitle("Model 1: proportion of births less than 2.5kg") + 
  theme_bw(base_size = 16) + 
  scale_color_manual(name = "", 
                     values = c("observed" = "darkblue"))+
  scale_fill_manual(name = "", 
                     values = c("replicated" = "lightblue")) 
```

```{r}
ggplot(data = as_tibble(t_y_rep_2), aes(value)) + 
    geom_histogram(aes(fill = "replicated")) + 
    geom_vline(aes(xintercept = t_y, color = "observed"), lwd = 1.5) + 
  ggtitle("Model 2: proportion of births less than 2.5kg") + 
  theme_bw(base_size = 16) + 
  scale_color_manual(name = "", 
                     values = c("observed" = "darkblue"))+
  scale_fill_manual(name = "", 
                     values = c("replicated" = "lightblue")) 
```

## Question 8

Based on the EDA, add sex as an additional covariates to the linear regression model. Then, the model become: $$\log(y_i) \sim N(\beta_1 + \beta_2\log(x_i) + \beta_3 s_i, \sigma^2)$$\
where\
- $y_i$ is weight in kg\
- $x_i$ is gestational age in weeks, CENTERED AND STANDARDIZED\
- $s_i$ is sex (0 for female and 1 for male)

```{r}
stan_data_mod3 <- list(N = nrow(ds),
                  log_weight = ds$log_weight,
                  log_gest = ds$log_gest_c,
                  sex = ifelse(ds$sex=="M",1,0))
mod3 <- stan(data = stan_data_mod3, 
             file ="Lab6mod3.stan",
             iter = 500,
             seed = 243)
```

```{r}
summary(mod3)$summary[c("beta[1]", "beta[2]","beta[3]", "sigma"),]
```

1.  Comparing the 100 simulated datasets drawn from the posterior predictive distribution to the actual data.

```{r}
yrep3 <- extract(mod3)[['log_weight_rep']]
samp3 <- sample(nrow(yrep3),100)
ppc_dens_overlay(y,yrep3[samp3,])
```

2.  Calculate the test statistic for the data, and the posterior predictive samples for both models, and plot the comparison.

```{r}
t_y_rep_3 <- sapply(1:nrow(yrep3), function(i) mean(yrep3[i,]<=log(2.5)))
ggplot(data = as_tibble(t_y_rep_3), aes(value)) + 
    geom_histogram(aes(fill = "replicated")) + 
    geom_vline(aes(xintercept = t_y, color = "observed"), lwd = 1.5) + 
  ggtitle("Model 3: proportion of births less than 2.5kg") + 
  theme_bw(base_size = 16) + 
  scale_color_manual(name = "", 
                     values = c("observed" = "darkblue"))+
  scale_fill_manual(name = "", 
                     values = c("replicated" = "lightblue")) 
```

Thus, the result shows that comparing to model 3, model 2 is better.
